---
title: Introducing Zoomerjoin
author: ''
date: '2023-07-19'
slug: []
draft: true
categories: []
tags: []
lastmod: '2023-07-19T23:10:23-04:00'
keywords: []
description: ''
comment: no
toc: no
autoCollapseToc: no
postMetaInFooter: no
hiddenFromHomePage: no
contentCopyright: no
reward: no
mathjax: yes
mathjaxEnableSingleDollar: yes
mathjaxEnableAutoNumber: no
hideHeaderAndFooter: no
flowchartDiagrams:
  enable: yes
  options: ''
sequenceDiagrams:
  enable: yes
  options: ''
bibliography: paper.bib
---

My new R package, Zoomerjoin, empowers you to fuzzily-join datasets at speeds you
never thought possible.

<!--more-->

<script type="text/javascript" id="MathJax-script" async
  src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js">
</script>

```{r, include=F}
library(tidyverse)
library(zoomerjoin)
knitr::opts_chunk$set(fig.path = "static")
set.seed(42)

n <- 500000
corpus_1 <- read_csv("bonica.csv")  %>%
    head(n)
names(corpus_1) <- c("id_1", "field")
corpus_2 <- read_csv("bonica.csv")  %>%
    tail(n)
names(corpus_2) <- c("id_2", "field")
```

If you work with large-scale social-science data, you are likely familiar with
the concept of "fuzzy" merging which describes a process to merge data that is
robust to minor misspellings or deviations between the keys used to identify
observations between datasets. Fuzzy merging allows you to merge data even when
fields have been inconsistently recorded or misspelled, as is common with
large-scale administrative data.

However, there's a catch: almost all popular fuzzy merging software packages
take *quadratic time,*  meaning they do not scale to large datasets. My new R
package, Zoomerjoin, empowers you to fuzzily join datasets in linear time,
allowing you to fuzzy join larger datasets than ever before.

# Background:

Fuzzy matching is typically taken to mean identifying all pairs of observations
between two datasets that have distance less than a specified threshold.
Existing fuzzy-joining methods in R do not scale to large datasets as they
exhaustively compare all possible pairs of units and recording all matching
pairs, incurring a quadratic $\mathcal{O}(mn)$ time cost.  Perhaps worse, the
most widely-used software packages typically also have a space complexity of
$O(mn)$, meaning that a patient user cannot simply wait for the join to
complete, as the memory of even large machines will be quickly exhausted [
@fuzzyjoin ].

Zoomerjoin solves this problem by implementing two Locality-Sensitive Hashing
algorithms [@Broder; @Datar_2004] which sort observations into buckets using a
bespoke hash function which assigns similar entries the same key with high
probability, while dissimilar items are unlikely to be assigned the same key.
After this initial sorting step, the algorithm checks pairs of records in the
same bucket to see if they are close enough to be considered a match. Records
in different buckets are never compared, so the algorithm takes
$O(\max_{ij}{(m_i n_j)})$ time to run (time proportional to the size of the
largest hash bucket). In the ordinary case that each observation matches to
few points in another dataset, the running time is dominated by the hashing
step, and the program finishes in linear time using linear memory.

# How to use

If you are familar with the dplyr-style logical joins (such as `inner_join`),
you are already familiar with the syntax of Zoomerjoin. Zoomerjoin supports
fuzzy joining for two distance measures, the Jaccard distance measure (for text
data), and the Euclidean distance (for points or vectors). You can perform a
fuzzy-join using these distance metrics, you can use `jaccard_(operation)_join`
or `euclidean_(operation)_join` respectively.

Here's an example showing how to merge two datasets using the
`jaccard_inner_join` function. I use as test datasets two random subsets
political donors listed in the [Database on Ideology, Money in Politics, and
Elections (DIME)](https://data.stanford.edu/dime) (the same test set from
benchmark from [this paper](https://www.cambridge.org/core/journals/political-analysis/article/adaptive-fuzzy-string-matching-how-to-merge-datasets-with-only-one-messy-identifying-field/275D7890548359215AC728C1E35B53CE)).

To give you a sense of the data, here's the first ten rows of the first sample:

```{r}
corpus_1
```

And the first ten rows of the second sample:

```{r}
corpus_2
```

I want to find matching rows or donors between the two tables, but even after
standardization, that the same donor may be spelled or recorded differently in
each dataset, so fuzzy joining is necessary.  I use the `jaccard_inner_join`
function to find matching rows between the two samples:

```{r}
start_time <- Sys.time()
join_out <- jaccard_inner_join(corpus_1, corpus_2,
                               by = "field",
                               threshold = .9, n_gram_width=6,
                               n_bands=40, band_width=6)
print(Sys.time() - start_time)
select(join_out, field.x, field.y) %>%
    mutate_all(~substr(.x,1,30))
```

The join finishes in just a few seconds on a modern data science laptop, and
uses very little memory, even though a brute-force comparison of all pairs
would involve $2.5^{11}$ operations! You can see a more detailed example of this
vignette and more of what you can do with zoomerjoin in the [introductory
vignette](https://beniaminogreen.github.io/zoomerjoin/articles/guided_tour.html),
and [package README](https://github.com/beniaminogreen/zoomerjoin).

# Benchmarks

How much time can zoomerjoin save you? The de facto standard for fuzzyjoining
is the venerable `fuzzyjoin` package which implements fuzzy joins for a panoply
of distance metrics. The package's most commonly used fuzzy joins, those for
the string distance are backed by optimized C code used to compare the strings.
However, `fuzzyjoin` runs an exhaustive search between all pairs of
observations between each dataset, it has to perform an increasing amount of
work for each additional row it seeks to merge, and hence does not scale to
large datasets.

Below, I compare the performance of `fuzzyjoin` to `zoomerjoin` when matching
datasets for the two distance metrics supported by `zoomerjoin`.

```{r, echo=F, message=F}
sim_data <- read_csv("sim_data.csv")
sim_data  %>%
    mutate(
           name = ifelse(name == "time", "Time Usage (s)", "Memory Usage (MB)"),
           join_type = ifelse(join_type == "Jaccard Distance",
                              "Jaccard Distance Join",
                              "Euclidean Distance Joins"),
           ) %>%
    ggplot(aes(x=as.numeric(n), y=value, col = package, linetype = package)) +
    xlab("Number of rows in each dataset") +
    geom_point() +
    geom_line() +
    theme_bw() +
    facet_wrap(~ join_type + name, scales = 'free') +
    scale_y_continuous("Time (s) / memory (MB)")
```

Zoomerjoin achieves almost linear scaling in both runtime and memory, while
`fuzzyjoin` scales quadratically. Even for datasets with 2500 rows,
`zoomerjoin` finishes in under a second. By contrast, the Jaccard-distance
joins implemented in `fuzzyjoin` take over three minutes to join. For the
largest Euclidean datasets, `fuzzyjoin` almost exhausts the 8GB memory capacity
of the laptop used for benchmarking, while `zoomerjoin`'s memory rises above
above 8 MB — a thousand-fold decrease.

Of course, `zoomerjoin` isn't better for every task; it supports fewer distance
metrics, as well as other types of joins such as regex joins, and joins for
genomic data.  Additionally, fuzzyjoin does not use Rust, so it is easier to
deploy on most systems, and will be available on more systems.

# Use Zoomerjoin For Blocking:

In many cases, a single distance metric will not be sufficient to distinguish
matches from non-matches. As an example, it might be that multiple fields are
needed to conclusively identify matches, or you might need to combine multiple
string-distance metrics to properly distinguish different spellings of the same
identifier from similar yet distinct identifiers.

Zoomerjoin can't do this itself, but it can drastically cut down on the time
needed to perform such matching by acting as a pre-processing step. To do this,
you can have Zoomerjoin return all pairs that are remotely similar, and then
distinguish between matches and non-matches using a bespoke model. I can't
scoop some of the work I'm doing currently, but I can share that I have used
Zoomerjoin as a pre-processing step to a machine-learning model to find
credible matches in datasets with hundreds of millions of rows in a matter of
minutes.

# Bibliography


