---
title: Creating R Plots without X11
author: ''
date: '2023-11-16'
slug: []
categories: []
tags: []
lastmod: '2023-11-16T17:08:48-05:00'
keywords: []
description: ''
comment: no
toc: no
autoCollapseToc: no
postMetaInFooter: no
hiddenFromHomePage: no
contentCopyright: no
reward: no
mathjax: no
mathjaxEnableSingleDollar: no
mathjaxEnableAutoNumber: no
hideHeaderAndFooter: no
flowchartDiagrams:
  enable: no
  options: ''
sequenceDiagrams:
  enable: no
  options: ''
---

Cypress pros don't want you to know about this one weird trick!

<!--more-->

```{r, echo=F, include=F}
library(tidyverse)
knitr::opts_chunk$set(fig.path = "static")
```

Every now and then, I'll have to work on a server where R has been compiled
without support for graphics. I thought this meant that I couldn't make plots,
but that's not the case! In this post, I'll share a way to render plots
in an R installation that's been compiled without graphics support
using the `Cario` library.

Normally, the easiest way around an R install without graphics support is to
run your compute on the server you're using, then download the results to your
local computer and make the figures there.  If you're like me and work with
restricted data, this likely isn't an option for you and you might have to get
inventive.

## How do you know if you have this issue?

The easiest way to diagnose if you have this issue is to use the
`capabilities()` command in R, which reports on the features R was compiled
With.  If you have `FALSE` under X11, jpeg, png, cairo, and tiff you are likely going
to have an issue.  In my case, I have `TRUE` under jpeg, png, and tiff, cairo, but
they aren't working because I don't have X11 in my environment.

```
capabilities()
jpeg         png        tiff       tcltk         X11        aqua
       TRUE        TRUE        TRUE        TRUE       FALSE       FALSE
   http/ftp     sockets      libxml        fifo      cledit       iconv
       TRUE        TRUE        TRUE        TRUE        TRUE        TRUE
        NLS       Rprof     profmem       cairo         ICU long.double
       TRUE        TRUE        TRUE        TRUE        TRUE        TRUE
    libcurl
       TRUE
```

## How To Resolve The Issue

I thought that this wasn't an issue that a regular user could fix, as it seems
like you would have to re-compile R to enable graphics support. As it turns
out, you can install the `Caro` package in R which provides another set of
graphics drivers and does not depend on X11, allowing you to render your graphs
even on a server without X11. Usage is very simple, but you have to manually
open and close each file you want to save graphics to, and I can't get the
`ggsave` command to work with it.

Here's how you can use `Cario` to render and save basic ggplot:

```{r, eval=F}
#install.packages(c("Cario","ggplot2"))
library(Cairo)
library(ggplot2)

Cairo(600, 600, file="basic_plot.png", type="png", bg="white") # open graphics device
ggplot(cars, aes(x=speed, y=dist)) + # plot a simple chart with ggplot
        geom_point() +
        ggtitle("Example Plot")
dev.off()
```

## Output:

A perfectly-rendered graph made without X11:

```{r, echo=F}
ggplot(cars, aes(x=speed, y=dist)) + # plot a simple chart with ggplot
        geom_point() +
        ggtitle("Example Plot")
```

